---
title: "LRRK2 DaTscan Analysis"
author: "Jackson Schumacher"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

check_and_load <- function(package, use_source = FALSE) {
  if (!require(package, character.only = TRUE)) {
    if (use_source) {
      install.packages(package, dependencies = TRUE, type = "source")
    } else {
      install.packages(package, dependencies = TRUE)
    }
    library(package, character.only = TRUE)
  }
}

setwd("/Users/jackson/Desktop/LRRK2 MS Materials")
```

```{r cars}
check_and_load("dplyr")
check_and_load("readxl")
check_and_load("tableone")
check_and_load("writexl")
check_and_load("tidyr")
check_and_load("VennDiagram")
check_and_load("grid")

check_and_load("lme4")
check_and_load("lmerTest")
check_and_load("ggplot2")
check_and_load("sjPlot")
check_and_load("broom.mixed")
check_and_load("patchwork")
check_and_load("ggpubr")
check_and_load("RColorBrewer")
```

# Baselinedemographics

## Data read in


```{r}
file_path <- "/Users/jackson/Desktop/LRRK2 MS Materials/LRRK2 PPMI (2023 V4) Analysis Data (Latest) (SAA Negative SNPs).xlsx"

data_dem1 <- read_excel(file_path, sheet = "Demographics (Latest)")

# Display structure of the data
str(data_dem1)

# Check first few rows to understand variable types
head(data_dem1)

# Show the number of unique values in data_dem$`Subject ID`
num_unique_subject_id <- data_dem1 %>%
  summarise(n_unique = n_distinct(`Subject ID`))
print(num_unique_subject_id)

# Check for missing values in all variables
na_summary <- data_dem1 %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)
```

```{r}
data_dem1 <- data_dem1  %>%
  mutate(across(where(is.character) & !all_of("Subject ID"), as.factor))

# Display the levels of all factor variables in data_dem
sapply(data_dem1, levels)
```

```{r}
# Reorder factor levels for the specified variables
data_dem <- data_dem1 %>%
  mutate(
    Cohort = factor(Cohort, levels = c("SPD", "G2019S PD", "M1646T R1441C/G PD", "N2081D/N14D PD")),
    `SAA Result` = factor(`SAA Result`, levels = c("Negative", "Positive", "Inconclusive", "Mixed Results", "Unknown")),
    Sex = factor(Sex, levels = c("Female", "Male")),
    Ethnicity = factor(Ethnicity, levels = c("Hispanic or Latino", "Not Hispanic or Latino", "Unknown")),
    Race = factor(Race, levels = c("White", "Black or African American", "Asian", "Native Hawaiian or Other Pacific Islander", "Other", "Multiracial", "Unknown")),
    `Years of Education` = factor(`Years of Education`, levels = c("0 years", "Less than 12 years", "12-16 years", "Greater than 16 years")),
    `Parent with PD` = factor(`Parent with PD`, levels = c("No", "Yes", "Unknown")),
    `Other Relative with PD` = recode(`Other Relative with PD`, "no" = "No"),
    `Other Relative with PD` = factor(`Other Relative with PD`, levels = c("No", "Yes", "Unknown"))
  )

# Display the reordered factor levels
sapply(data_dem, levels)

# Show the number of unique values in data_dem$`Subject ID`
num_unique_subject_id <- data_dem %>%
  summarise(n_unique = n_distinct(`Subject ID`))
print(num_unique_subject_id)

na_summary <- data_dem %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)
```

Use data_dem as the main analysis dataframe for future analysis.

```{r, eval=FALSE}
save(data_dem, file = "data_dem.RData")
```
```{r}
file_path <- "/Users/jackson/Desktop/LRRK2 MS Materials/LRRK2 PPMI (2023 V4) Analysis Data (Latest) (SAA Negative SNPs).xlsx"

data_datscan1 <- read_excel(file_path, sheet = "DaTscan")

# Display structure of the data
str(data_datscan1)

# Check first few rows to understand variable types
head(data_datscan1)

# Check for missing values in all variables
na_summary <- data_datscan1 %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)

# Remove observations with NA
data_datscan2 <- data_datscan1 %>%
  filter(!is.na(visit_year)) 
```

```{r}
data_datscan2 <- data_datscan2 %>%
  mutate(across(where(is.character) & !all_of('participant_id'), as.factor))

sapply(data_updrs2, levels)
```

```{r}
data_datscan3 <- data_datscan2 %>%
  separate_rows(`Group ID`, sep = " \\| ") %>%
  mutate(
    `Group ID` = factor(`Group ID`, levels = c("SPD", "G2019S PD", "M1646T R1441C/G PD", "N2081D/N14D PD")),
    sbr_caudate = (sbr_caudate_r + sbr_caudate_l) / 2,
    sbr_putamen = (sbr_putamen_r + sbr_putamen_l) / 2
  )

sapply(data_datscan3, levels)
```

```{r}
# Select the columns to merge from data_dem
data_dem_selected <- data_dem %>%
  select(`Subject ID`, `Age at baseline`, Sex, Race, Ethnicity, `Years of Education`, `Parent with PD`, `Other Relative with PD`) %>%
  mutate(
    Race = factor(if_else(Race == "White", "White", "Other"), levels = c("White", "Other")),
    Ethnicity = factor(if_else(Ethnicity == "Hispanic or Latino", "Hispanic or Latino", "Other"), levels = c("Hispanic or Latino", "Other")),
    `Years of Education` = factor(if_else(`Years of Education` == "0 years", "Less than 12 years", `Years of Education`), levels = c("Less than 12 years", "12-16 years", "Greater than 16 years"))
  ) %>%
  group_by(`Subject ID`) %>%
  summarise(across(everything(), first))  # Take the first occurrence for each `Subject ID`

# Merge the selected columns into data_datscan
data_datscan3 <- data_datscan3 %>%
  left_join(data_dem_selected, by = c("participant_id" = "Subject ID"))

data_updrs_selected <- data_updrs %>%
  select(participant_id, upd23a_medication_for_pd) %>%
  group_by(participant_id) %>%
  slice(1) %>%
  ungroup()

data_datscan3 <- data_datscan3 %>%
  left_join(data_updrs_selected, by = "participant_id")

# Extract baseline values for each outcome score
baseline_values <- data_datscan3 %>%
  group_by(participant_id) %>%
  filter(visit_year == min(visit_year)) %>%
  slice(1) %>%
  select(participant_id, 
         baseline_sbr_caudate_r = sbr_caudate_r, 
         baseline_sbr_caudate_l = sbr_caudate_l, 
         baseline_sbr_caudate = sbr_caudate, 
         baseline_sbr_putamen_r = sbr_putamen_r, 
         baseline_sbr_putamen_l = sbr_putamen_l, 
         baseline_sbr_putamen = sbr_putamen)%>%
  ungroup()

# Merge the baseline values into the main data frame
data_datscan <- data_datscan3 %>%
  left_join(baseline_values, by = "participant_id")

# Display the first few rows to verify the changes
str(data_datscan)
```

```{r}
# Check for missing values in all variables
na_summary <- data_datscan %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)
```

Show the number of unique Subject ID

```{r}
num_unique_subject_id <- data_datscan %>%
  summarise(n_unique = n_distinct(participant_id))
print(num_unique_subject_id)
```

Count the number of records for each participant_id

```{r}
participant_counts <- data_datscan %>%
  count(participant_id) %>%
  rename(number_of_records = n)

# Display the first few rows of the counts
head(participant_counts)

# Plot the distribution of the number of repeat measures
ggplot(participant_counts, aes(x = number_of_records)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Number of Repeat Measures per Participant",
       x = "Number of Repeat Measures",
       y = "Count of Participants") +
  theme_classic()
```

Create a histogram to show the distribution of visit_year

```{r}
histogram_visit_year <- ggplot(data_datscan, aes(x = visit_year)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "black", alpha = 0.7) +
  theme_classic() +
  labs(
    title = "Distribution of Visit Month",
    x = "Visit Month",
    y = "Count"
  )+
  scale_y_continuous(limits = c(0, 1000)) +  # Set y-axis range
  scale_x_continuous(limits = c(-1, 6))  # Set x-axis range


# Display the histogram
histogram_visit_year
```

## Table S4: Table 1 for DaTscan

```{r}
data_datscan_dem <- data_datscan %>% 
  group_by(participant_id, `Group ID`) %>%
  filter(visit_year == min(visit_year)) %>%
  slice(1) %>%
  ungroup()

# Define variables and categorical variables
vars <- c("Age at baseline", "Sex", "Race", "Ethnicity", "Years of Education", "Parent with PD", "Other Relative with PD", "GBA Mutation", "SAA Result")
catVars <- c("Sex", "Race", "Ethnicity", "Years of Education", "Parent with PD", "Other Relative with PD", "GBA Mutation", "SAA Result")

# Generate Table 1
table1 <- CreateTableOne(vars = vars, strata = "Group ID", data = data_datscan_dem, factorVars = catVars)

```

# 1. LRRK2 and DaTscan 

## Model 1 Caudate: Unadjusted

```{r}
# Aggregate the data by visit_name and Group ID
data_datscan_summary <- data_datscan %>%
  group_by(visit_year, `Group ID`) %>%
  summarise(
    caudate_mean_score = mean(sbr_caudate, na.rm = TRUE),
    caudate_sd_score = sd(sbr_caudate, na.rm = TRUE),
    putamen_mean_score = mean(sbr_putamen, na.rm = TRUE),
    putamen_sd_score = sd(sbr_putamen, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(caudate_se_score = caudate_sd_score / sqrt(n),
         putamen_se_score = putamen_sd_score / sqrt(n))  # Calculate standard error
```

```{r}
caudate_model1 <- lmer(
  sbr_caudate ~ visit_year * `Group ID` + baseline_sbr_caudate + (1 | participant_id),
  data = data_datscan
)

tab_model(caudate_model1, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, title = "Caudate")

plot_caudate_model1 <- ggplot(data_datscan_summary, aes(x = visit_year, y = caudate_mean_score, color = `Group ID`, group = `Group ID`)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = caudate_mean_score - caudate_se_score, ymax = caudate_mean_score + caudate_se_score), width = 0.5) +
  theme_classic() + 
  labs(
    x = "Visit month",
    y = "Caudate",
    color = "LRRK2 SNP"
  ) + 
  theme(
    legend.position = "bottom",
    plot.title = element_blank()
  ) + 
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(1, 2.5)) +  # Set y-axis range
  scale_x_continuous(limits = c(0, 5))  # Set x-axis range

plot_caudate_model1

legend <- ggpubr::get_legend(plot_caudate_model1)
```

## Model 1 Putamen 

```{r}
putamen_model1 <- lmer(
  sbr_putamen ~ visit_year * `Group ID` + baseline_sbr_putamen + (1 | participant_id),
  data = data_datscan
)

tab_model(putamen_model1, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, title = "Putamen")

plot_putamen_model1 <- ggplot(data_datscan_summary, aes(x = visit_year, y = putamen_mean_score, color = `Group ID`, group = `Group ID`)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = putamen_mean_score - putamen_se_score, ymax = putamen_mean_score + putamen_se_score), width = 0.5) +
  theme_classic() + 
  labs(
    x = "Visit month",
    y = "Putamen",
    color = "LRRK2 SNP"
  ) + 
  theme(
    legend.position = "bottom",
    plot.title = element_blank()
  ) +
  scale_y_continuous(limits = c(0, 1.2)) +  # Set y-axis range
  scale_x_continuous(limits = c(0, 5))  # Set x-axis range

plot_putamen_model1
```

## Model 2 Caudate: Adjusted for "Age at baseline", "Sex", "Race", "Ethnicity", "Years of Education", "upd23a_medication_for_pd"

```{r}
# Fit a linear mixed-effects model
caudate_model2 <- lmer(
  sbr_caudate ~ visit_year * `Group ID` + baseline_sbr_caudate + `Age at baseline` + Sex + Race + Ethnicity + `Years of Education` + upd23a_medication_for_pd + (1 | participant_id),
  data = data_datscan
)

tab_model(caudate_model2, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, title = "Caudate")

# Visualize the fixed effects of the model
plot_caudate_model2 <- sjPlot::plot_model(caudate_model2, type = "pred", ci.lvl = NA, terms = c("visit_year", "Group ID"), color = "breakfast club") + 
  geom_line(linewidth = 0.25) + 
  theme_classic() + 
  labs(
    x = "Visit year",
    y = "Caudate",
    color = "LRRK2 SNP"
  ) + 
  theme(
    legend.position = "bottom",
    plot.title = element_blank()
  ) +
  scale_y_continuous(limits = c(1.25, 2)) +  # Set y-axis range
  scale_x_continuous(limits = c(0, 4))  # Set x-axis range

plot_caudate_model2
```

## Model 2 Putamen

```{r}
# Fit a linear mixed-effects model
putamen_model2 <- lmer(
  sbr_putamen ~ visit_year * `Group ID` + baseline_sbr_putamen + `Age at baseline` + Sex + Race + Ethnicity + `Years of Education` + upd23a_medication_for_pd + (1 | participant_id),
  data = data_datscan
)

tab_model(putamen_model2, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, title = "Putamen")

# Visualize the fixed effects of the model
plot_putamen_model2 <- sjPlot::plot_model(putamen_model2, type = "pred", ci.lvl = NA, terms = c("visit_year", "Group ID"), color = "breakfast club") + 
  geom_line(linewidth = 0.25) +
  theme_classic() + 
  labs(
    x = "Visit month",
    y = "Putamen",
    color = "LRRK2 SNP"
  ) + 
  theme(
    legend.position = "bottom",
    plot.title = element_blank()
  ) +
  scale_y_continuous(limits = c(0.5, 0.9)) +  # Set y-axis range
  scale_x_continuous(limits = c(0, 4))  # Set x-axis range

plot_putamen_model2
```
