---
title: "LRRK2 Overall Analysis"
author: "Jackson Schumacher"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

check_and_load <- function(package, use_source = FALSE) {
  if (!require(package, character.only = TRUE)) {
    if (use_source) {
      install.packages(package, dependencies = TRUE, type = "source")
    } else {
      install.packages(package, dependencies = TRUE)
    }
    library(package, character.only = TRUE)
  }
}

setwd("/Users/jackson/Desktop/LRRK2 MS Materials")
```

```{r check_and_load}
check_and_load("dplyr")
check_and_load("readxl")
check_and_load("tableone")
check_and_load("writexl")
check_and_load("tidyr")
check_and_load("VennDiagram")
check_and_load("grid")

check_and_load("lme4")
check_and_load("lmerTest")
check_and_load("ggplot2")
check_and_load("sjPlot")
check_and_load("broom.mixed")
check_and_load("patchwork")
check_and_load("ggpubr")
check_and_load("RColorBrewer")
```

# Baseline demographics

## Data read in

```{r}
file_path <- "/Users/jackson/Desktop/LRRK2 MS Materials/LRRK2 PPMI (2023 V4) Analysis Data (Latest) (SAA Positive).xlsx"

data_dem1 <- read_excel(file_path, sheet = "Demographics (Latest)")

# Display structure of the data
str(data_dem1)

# Check first few rows to understand variable types
head(data_dem1)

# Show the number of unique values in data_dem$`Subject ID`
num_unique_subject_id <- data_dem1 %>%
  summarise(n_unique = n_distinct(`Subject ID`))
print(num_unique_subject_id)

# Check for missing values in all variables
na_summary <- data_dem1 %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)
```

```{r}
data_dem1 <- data_dem1  %>%
  mutate(across(where(is.character) & !all_of("Subject ID"), as.factor))

# Display the levels of all factor variables in data_dem
sapply(data_dem1, levels)
```

```{r}
# Reorder factor levels for the specified variables
data_dem <- data_dem1 %>%
  mutate(
    Cohort = factor(Cohort, levels = c("SPD", "LRRK2 PD", "GBA PD", "LRRK2-GBA PD")),
    `SAA Result` = factor(`SAA Result`, levels = c("Negative", "Positive", "Inconclusive", "Mixed Results", "Unknown")),
    Sex = factor(Sex, levels = c("Female", "Male")),
    Ethnicity = factor(Ethnicity, levels = c("Hispanic or Latino", "Not Hispanic or Latino", "Unknown")),
    Race = factor(Race, levels = c("White", "Black or African American", "Asian", "Native Hawaiian or Other Pacific Islander", "Other", "Multiracial", "Unknown")),
    `Years of Education` = factor(`Years of Education`, levels = c("0 years", "Less than 12 years", "12-16 years", "Greater than 16 years")),
    `Parent with PD` = factor(`Parent with PD`, levels = c("No", "Yes", "Unknown")),
    `Other Relative with PD` = recode(`Other Relative with PD`, "no" = "No"),
    `Other Relative with PD` = factor(`Other Relative with PD`, levels = c("No", "Yes", "Unknown"))
  )

# Display the reordered factor levels
sapply(data_dem, levels)

# Show the number of unique values in data_dem$`Subject ID`
num_unique_subject_id <- data_dem %>%
  summarise(n_unique = n_distinct(`Subject ID`))
print(num_unique_subject_id)

na_summary <- data_dem %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)
```

Use data_dem as the main analysis dataframe for future analysis.

```{r, eval=FALSE}
save(data_dem, file = "data_dem.RData")
```

# 1. LRRK2 and UPDRS analysis

## Data read in
Changes made from the original data received: particpant_id renamed to participant_id

```{r}
data_updrs1 <- read_excel(file_path, sheet = "UPDRS III")

str(data_updrs1)
head(data_updrs1)

na_summary <- data_updrs1 %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)

data_updrs1 <- data_updrs1 %>%
  mutate(across(where(is.character) & !all_of('participant_id'), as.factor))

sapply(data_updrs1, levels)
```

```{r}
data_updrs2 <- data_updrs1 %>%
  separate_rows(`Group ID`, sep = " \\| ") %>%
  mutate(
    `Group ID` = factor(`Group ID`, levels = c("SPD", "LRRK2 PD", "GBA PD", "LRRK2-GBA PD")),
    upd23a_medication_for_pd = factor(upd23a_medication_for_pd, levels = c("No", "Yes", "Unknown")),
    upd23b_clinical_state_on_medication = if_else(is.na(upd23b_clinical_state_on_medication), "Unknown", upd23b_clinical_state_on_medication),
    upd23b_clinical_state_on_medication = factor(upd23b_clinical_state_on_medication)
    )

sapply(data_updrs2, levels)
```

```{r}
# Select the columns to merge from data_dem
data_dem_selected <- data_dem %>%
  select(`Subject ID`, `Age at baseline`, Sex, Race, Ethnicity, `Years of Education`) %>%
  mutate(
    Race = factor(if_else(Race == "White", "White", "Other"), levels = c("White", "Other")),
    Ethnicity = factor(if_else(Ethnicity == "Hispanic or Latino", "Hispanic or Latino", "Other"), levels = c("Hispanic or Latino", "Other")),
    `Years of Education` = factor(if_else(`Years of Education` == "0 years", "Less than 12 years", `Years of Education`), levels = c("Less than 12 years", "12-16 years", "Greater than 16 years"))
  ) %>%
  group_by(`Subject ID`) %>%
  summarise(across(everything(), first))  # Take the first occurrence for each `Subject ID`

sapply(data_dem_selected, levels)

# Merge the selected columns into data_updrs
data_updrs3 <- data_updrs2 %>%
  left_join(data_dem_selected, by = c("participant_id" = "Subject ID"))

# Display the merged data
str(data_updrs3)
```

```{r}
# Ensure unique baseline measures for each participant
baseline_measures <- data_updrs3 %>%
  arrange(participant_id, visit_year) %>%
  group_by(participant_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(participant_id, baseline_score = mds_updrs_part_iii_summary_score)

# Merge the baseline measure into the main data frame
data_updrs <- data_updrs3 %>%
  left_join(baseline_measures, by = "participant_id")

# Check for missing values in all variables
na_summary <- data_updrs %>%
  summarise_all(~ sum(is.na(.)))%>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "MissingValues")
print(na_summary)
```

```{r}
# Count the number of records for each participant_id
participant_counts <- data_updrs %>%
  count(participant_id) %>%
  rename(number_of_records = n)

# Display the first few rows of the counts
head(participant_counts)

# Plot the distribution of the number of repeat measures
ggplot(participant_counts, aes(x = number_of_records)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Number of Repeat Measures per Participant",
       x = "Number of Repeat Measures",
       y = "Count of Participants") +
  theme_classic()

# Count the number of participants with exactly one record
n_single_record <- participant_counts %>%
  filter(number_of_records == 1) %>%
  summarise(N = n())

print(n_single_record)
```

```{r}
# Calculate summary statistics
summary_stats <- data_updrs %>%
  summarise(
    mean = mean(visit_year, na.rm = TRUE),
    sd = sd(visit_year, na.rm = TRUE),
    median = median(visit_year, na.rm = TRUE),
    min = min(visit_year, na.rm = TRUE),
    max = max(visit_year, na.rm = TRUE)
  )

# Print summary statistics
print(summary_stats)

# Create a histogram to show the distribution of visit_year
histogram_visit_year <- ggplot(data_updrs, aes(x = visit_year)) +
  geom_histogram(binwidth = 1, fill = "yellowgreen", color = "black", alpha = 0.7) +
  theme_classic() +
  labs(
    title = "Distribution of Visit Year",
    x = "Visit Year",
    y = "Count"
  ) +
  scale_y_continuous(limits = c(0, 2500)) +  # Set y-axis range
  scale_x_continuous(limits = c(-1, 15))  # Set x-axis range

# Display the histogram
histogram_visit_year
```

Use data_updrs as the main analysis dataframe for future analysis.

```{r, eval=FALSE}
save(data_updrs, file = "data_updrs.RData")
```

## Model 1: Unadjusted

```{r}
# Fit a linear mixed-effects model
updrs_model1 <- lmer(
  mds_updrs_part_iii_summary_score ~ visit_year * `Group ID` + baseline_score + (1 | participant_id),
  data = data_updrs
)

# Summary of the model
summary(updrs_model1)

# Create a publishable table using sjPlot
tab_model(updrs_model1, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, 
          title = "Model Coefficients and Slopes for Time Effect by Groups")
```

```{r}
# Filter out participants with only one measurement
filtered_data <- data_updrs %>%
  group_by(participant_id) %>%
  filter(n() > 1) %>%
  ungroup()

# Fit the linear mixed-effects model on filtered data
updrs_model_filtered <- lmer(
  mds_updrs_part_iii_summary_score ~ visit_year * `Group ID` + baseline_score + (1 | participant_id),
  data = filtered_data
)

tab_model(updrs_model_filtered, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, 
          title = "Model Coefficients and Slopes for Time Effect by Groups")
```

```{r}
# Categorize visit_year into bins of every 12 months
data_updrs <- data_updrs %>%
  mutate(visit_bin = cut(visit_year, breaks = seq(0, max(visit_year, na.rm = TRUE), by = 1), include.lowest = TRUE, right = FALSE))

# Aggregate the data by visit_bin and Group ID
data_updrs_summary <- data_updrs %>%
  group_by(visit_bin, `Group ID`) %>%
  summarise(
    mean_score = mean(mds_updrs_part_iii_summary_score, na.rm = TRUE),
    sd_score = sd(mds_updrs_part_iii_summary_score, na.rm = TRUE),
    n = n(),
    visit_year_median = median(visit_year, na.rm = TRUE)
  ) %>%
  mutate(se_score = sd_score / sqrt(n))  # Calculate standard error

# Display the summary data
head(data_updrs_summary)

# Create the plot
plot_updrs_model1 <- ggplot(data_updrs_summary, aes(x = visit_year_median, y = mean_score, color = `Group ID`, group = `Group ID`)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_score - se_score, ymax = mean_score + se_score), width = 0.5) +
  theme_classic() + 
  labs(
    x = "Visit year",
    y = "MDS-UPDRS Part III score",
    color = "LRRK2 SNP"
  ) + 
  theme(
    legend.position = "bottom",
    plot.title = element_blank()
  ) + 
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(9, 54)) +  # Set y-axis range
  scale_x_continuous(limits = c(0, 14))  # Set x-axis range

plot_updrs_model1

# Extract the legend
legend <- ggpubr::get_legend(plot_updrs_model1)

# Display the legend
ggpubr::as_ggplot(legend)
```

## Model 2: Adjusted for "Age at baseline", "Sex", "Race", "Ethnicity", "Years of Education", "upd23a_medication_for_pd"

Multivariable adjusted progression of MDS-UPDRS Part III score in Parkinson’s Disease patients carrying different LRRK2 mutations.

```{r}
# Fit a linear mixed-effects model
updrs_model2 <- lmer(
  mds_updrs_part_iii_summary_score ~ visit_year * `Group ID` + baseline_score + `Age at baseline` + Sex + Race + Ethnicity + `Years of Education` + upd23a_medication_for_pd + (1 | participant_id),
  data = data_updrs
)

# Create a publishable table using sjPlot
tab_model(updrs_model2, show.re.var = FALSE, show.icc = FALSE, show.r2 = FALSE, 
          title = "Model Coefficients and Slopes for Time Effect by Groups")
```


```{r}
# Visualize the fixed effects of the model
plot_updrs_model2 <- sjPlot::plot_model(updrs_model2, type = "pred", ci.lvl = NA, terms = c("visit_year", "Group ID"), colors = "Set1") +
  geom_line(linewidth = 0.25) +
  theme_classic() +
  labs(
    x = "Years since baseline visit",
    y = "Mean UPDRS III score",
    color = "Group",
  ) + 
  theme(
    legend.position = "bottom",
    plot.title = element_blank(),
  ) +
  scale_y_continuous(limits = c(20, 50)) + # Set y-axis range 
  scale_x_continuous(limits = c(0, 15))  # Set x-axis range

plot_updrs_model2
```
